---
title: "Reproducible Research Course Project 1"
author: "Sayan Banerjee"
date: "Saturday, July 18, 2015"
output: html_document
---
Loading the required packages for the data analysis

```{r loadpackages, echo=TRUE, results="hide"}

require(dplyr)
require(knitr)
require(xtable)
require(ggplot2)
```

```{r setopions, echo=FALSE}
opts_chunk$set(echo=TRUE, results="asis")
```


###Data
The data for this assignment can be downloaded from the course web site:
.Dataset: Activity monitoring data 

The variables included in this dataset are:

.steps: Number of steps taking in a 5-minute interval (missing values are coded as  NA )


.date: The date on which the measurement was taken in YYYY-MM-DD format


.interval: Identifier for the 5-minute interval in which measurement was taken


The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

The first step is to read the data and then to clean the rows with NA values for first set of analysis.

```{r readdata}
actDF <- read.csv("activity.csv", stringsAsFactors = FALSE)
actclDF <- filter(actDF,!is.na(steps)&!is.na(date)&!is.na(interval))

```

###Analysis Part 1 - What is mean total number of steps taken per day?

The histogram will display the total number of steps taken each day


```{r histsteps}
histDF <- actclDF %>% 
  group_by(date) %>%
  summarise(steps =sum(steps))

hist(histDF$steps, xlab="Steps per day")


```

The below presents the mean and median total number of steps taken per day

```{r statsteps}
statStepsDF <- actclDF %>% 
  group_by(date) %>%
  summarise(meanSteps =mean(steps), medianSteps = median(steps), totalSteps=sum(steps))

tab<- xtable(statStepsDF)
print(tab, type="html")
```

###Analysis Part 2 - What is the average daily activity pattern?

The below time series plot  displays 5-minute interval on x-axis and the average number of steps taken, averaged across all days on y-axis


```{r stepinterval}

intStepsDF <- actclDF %>% 
  group_by(interval) %>%
  summarise(avgSteps =mean(steps))

g <- ggplot(data=intStepsDF, aes(x=interval,y=avgSteps))+
  geom_line()

maxAvgStepInt <- filter(intStepsDF,avgSteps==max(avgSteps))$interval

g+geom_vline(xintercept=maxAvgStepInt, linetype="dotted")

```

The `r maxAvgStepInt`th 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.

###Analysis Part 3 - Imputing missing values

There are a number of days/intervals where there are missing values (coded as  NA ). The presence of missing days may introduce bias into some calculations or summaries of the data.

1st step is to calculate report the total number of missing values in the dataset (i.e. the total number of rows with  NA s)


```{r numNA}

numNA <- nrow(filter(actDF,is.na(steps)))

```

There are `r numNA` rows with NA s.

Next step  is to filling in all of the missing values in the dataset with mean for that 5-minute interval.

```{r impMissVal}

actImpDF<-actDF
for(i in 1:nrow(actImpDF))
  {
    if (is.na(actImpDF[i,"steps"]))
    {
      actImpDF[i,"steps"] <- round(filter(intStepsDF,interval==actImpDF[i,"interval"])$avgSteps)
    }
  }

```


The below histogram displays the total number of steps taken each day after imputing missing data.


```{r histimpSteps}
histImpDF <- actImpDF %>% 
  group_by(date) %>%
  summarise(steps =sum(steps))

hist(histImpDF$steps, xlab="Steps per day")


```


The below table report the mean and median total number of steps taken per day after imputing missing data.

```{r statimpsteps}
statImpStepsDF <- actImpDF %>% 
  group_by(date) %>%
  summarise(meanSteps =mean(steps), medianSteps = median(steps), totalSteps=sum(steps))

tab<- xtable(statImpStepsDF)
print(tab, type="html")
```

###Analysis Part 4 - Are there differences in activity patterns between weekdays and weekends?

1st step is to create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r weekday}

wday <- function(y)
         {
            x<-""
            if((weekdays(as.Date(y))=="Saturday") | (weekdays(as.Date(y))== "Sunday"))
             {
                x<-"weekend"
             }
           else
             {
               x<-"weekday"
             }
           x
         }
                    

day <- data.frame(sapply(actImpDF$date,wday))
actImpWDDF <- cbind(actImpDF,day)
names(actImpWDDF)[4]<-paste("wday") 

intWDStepsDF <- actImpWDDF %>% 
  group_by(interval,wday) %>%
  summarise(avgSteps =mean(steps))
```

The below panel plot shows a time series plot of the 5-minute interval on x-axi and the average number of steps taken, averaged across all weekday days or weekend days on y-axis.

```{r wdayplot}

ggplot(data=intWDStepsDF, aes(x=interval,y=avgSteps))+
  geom_line()+facet_grid(wday~.)

```